---
title: "New York Shooting Incidents Data Analysis"
author: "Marco Perez"
date: "2025-03-30"
output:
  html_document: default
  pdf_document: default
  Execute below command within Rstudio terminal:
    install.packages(c("forecast","ggplot2","zoo"))
---

## Introduction

Analyzing police department shooting incident data provides a detailed vast amount of vital information that can be utilized for the future. With both communities and the police department understanding specific data trends in their communities, good could be done with the data sets.New York provides a great set of diverse data between the population and different communities. 

## Import Data

```{r setup, echo= TRUE, include=FALSE}
library(tidyverse)
library(zoo)
library(forecast)
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
data<- read.csv(url)
```
## Tidy Data - Remove Unwanted Column Variables

```{r data-cleaning, include=TRUE}
clean_data <- data %>%
  select(-Longitude,-Latitude,-INCIDENT_KEY,-STATISTICAL_MURDER_FLAG,
         -X_COORD_CD,-Y_COORD_CD,-Lon_Lat,-JURISDICTION_CODE,-PRECINCT)

```

## Tidy Data - Combining like catagories with Variable

```{r data-catagory-combining, include=TRUE}
clean_data <- clean_data %>%
  mutate(
    PERP_SEX = case_when(
    PERP_SEX %in% c("", "(null)", "U") ~ "Unknown",
    TRUE ~ as.character(PERP_SEX)
  ),
  LOCATION_DESC = case_when(
    LOCATION_DESC %in% c("","(null)","Other") ~ "Other",
    TRUE ~ as.character(LOCATION_DESC)
  ),
  LOC_CLASSFCTN_DESC = case_when(
    LOC_CLASSFCTN_DESC %in% c("","OTHER","Other") ~ "Other",
    TRUE ~ as.character(LOC_CLASSFCTN_DESC)
  ),
  PERP_AGE_GROUP = case_when(
    PERP_AGE_GROUP %in% c("","(null)","UNKNOWN") ~ "Unknown",
    TRUE ~ as.character(PERP_AGE_GROUP)
  ),
  PERP_RACE = case_when(
    PERP_RACE %in% c("","(null)","UNKNOWN") ~ "Unkonwn",
    TRUE ~ as.character(PERP_RACE)
  ),
  VIC_AGE_GROUP = case_when(
    VIC_AGE_GROUP %in% c("1022","UNKNOWN") ~ "Unknown",
    TRUE ~ as.character(VIC_AGE_GROUP)
  ),
  LOC_OF_OCCUR_DESC = case_when(
    LOC_OF_OCCUR_DESC %in% c("") ~ "Unknown",
    TRUE ~ as.character(LOC_OF_OCCUR_DESC)
  )
  )

```

## Tidy Data - Correct Data Types

```{r data-catagoizing, include=TRUE}
library(lubridate)
clean_data$OCCUR_DATE <- mdy(clean_data$OCCUR_DATE)
clean_data$OCCUR_TIME <- hms(clean_data$OCCUR_TIME)
clean_data$BORO <- as.factor(clean_data$BORO)
clean_data$LOC_OF_OCCUR_DESC <- as.factor(clean_data$LOC_OF_OCCUR_DESC)
clean_data$LOC_CLASSFCTN_DESC <- as.factor(clean_data$LOC_CLASSFCTN_DESC)
clean_data$LOCATION_DESC <- as.factor(clean_data$LOCATION_DESC)
clean_data$PERP_AGE_GROUP <- as.factor(clean_data$PERP_AGE_GROUP)
clean_data$PERP_SEX <- as.factor(clean_data$PERP_SEX)
clean_data$PERP_RACE <- as.factor(clean_data$PERP_RACE)
clean_data$VIC_AGE_GROUP <- as.factor(clean_data$VIC_AGE_GROUP)
clean_data$VIC_SEX <- as.factor(clean_data$VIC_SEX)
clean_data$VIC_RACE <- as.factor(clean_data$VIC_RACE)
summary(clean_data)
```

## Visualizing the Data

## Annual Trend related to number of Daylight Hours

```{r aggregate incident data, echo=FALSE}
incidents_by_day <- clean_data %>%
  group_by(OCCUR_DATE) %>%
  summarize(incidents = n())
incidents_by_day$smoothed_incidents <- rollmean(incidents_by_day$incidents, k = 60, fill = NA, align = "center")
ggplot(incidents_by_day, aes(x = OCCUR_DATE, y = smoothed_incidents)) +
     geom_line(color = "blue", alpha = 0.5) +
     geom_line(aes(y = incidents), color = "red", alpha = 0.5) +
     labs(title = "Shooting Incidents Over Time",
          x = "Date",
          y = "Number of Incidents",
          caption = "Red: Daily Incidents, Blue: 365-day Moving Average") +
     theme_minimal()

incidents_by_day_ts <- ts(incidents_by_day$incidents, frequency = 365)
incidents_decomposed <- decompose(incidents_by_day_ts)
plot(incidents_decomposed)
fit <- stlf(incidents_decomposed$seasonal)
future_periods <- 365
forecasted <- forecast(fit, h=future_periods)
plot(forecasted)
```


```{r model_analysis, echo=FALSE}
incidents_by_day$day_num <- as.numeric(incidents_by_day$OCCUR_DATE - min(incidents_by_day$OCCUR_DATE))
linear_model <- lm(incidents ~ day_num, data=incidents_by_day)
summary(linear_model)
ggplot(incidents_by_day, aes(x=day_num, y=incidents)) +
  geom_point(aes(color="Incidents")) + 
  geom_smooth(method="lm", se=FALSE, color="red", aes(color="Trend Line")) +
  labs(title="Shooting Incidents Over Time", color="Legend") +
  xlab("Days Since Start") +
  ylab("Number of Incidents")
```

## Conclusion

The frequency of shooting incidents exhibits a seasonal pattern, though the underlying cause remains unclear. Potential explanation include increased daylight hours during summer, leading to great public activity later in the day, or the inlfuence of high temeratures on agression. It is important to acknowledge potential biases in interpreting this data. Further investigation of this seasonal trend and other portential correlations is warranted. 


